---
title: "TimeSeries Prediction of Cancer in Japan"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Time series analysis using ARIMA model to cancer data

All data refer to sources [below](https://ganjoho.jp/reg_stat/statistics/data/dl/index.html).

- Cancer Statistics. Cancer Information Service, National Cancer Center, Japan (Vital Statistics of Japan, Ministry of Health, Labour and Welfare)  
- Cancer Statistics. Cancer Information Service, National Cancer Center, Japan　(Monitoring of Cancer Incidence in Japan (MCIJ))  
Cancer incidence and incidence rates in Japan in 2009: a study of 32 population-based cancer registries for the Monitoring of Cancer Incidence in Japan (MCIJ) project.[Japanese Journal of Clinical Oncology 2015; 45: 884-91](https://ganjoho.jp/external.html?url=https%3A%2F%2Fpubmed.ncbi.nlm.nih.gov%2F26142437%2F&title=Japanese%20Journal%20of%20Clinical%20Oncology%202015;%2045:%20884-91)  

In this Analysis some data include Japanese character (日本語) but we will try to refrain from using Japanese language data to avoid encoding problems.

```{r library, include=F, warning=F, message=F}
library(forecast)
library(tseries)
library(ggfortify)
library(lubridate)
library(tidyverse)
library(patchwork)
```

```{r,warning=F, message=F}
d_incidence <- read.csv("input/cancer_incidence_jpn.csv",check.names = F)
d_mortality <- read.csv("input/cancer_mortality_jpn.csv",check.names = F)
pop <- read.csv("input/population_all_jpn.csv",check.names = F)

head(d_incidence)
```

```{r,warning=F, message=F}
head(d_mortality)
```

```{r,warning=F, message=F}
head(pop)
```

Description:  
- Duration of incidence data is 1975 to 2015.  
- Duration of mortality data and population is 1958 to 2020.  
Column names:  
- "all" means all age count, other column names refer to each age range.  
- "part" means parts on the body where cancer happens. They are written in Japanese  
- "code" uniquely matches with "parts". For instance, 1 is all parts and 8 is liver.   

### data processing

To make analysis easier, change data shape wide to long.  

```{r warning=F, message=F}
d_incidence_long <- pivot_longer(d_incidence,cols=c("all","0_4","5_9","10_14","15_19",         
                                "20_24","25_29","30_34","35_39","40_44",         
                                "45_49","50_54","55_59","60_64","65_69",         
                                "70_74","75_79","80_84","over85"),
                       names_to = "age",
                       values_to = "count") %>% 
  filter(code %in% c(1:67)) %>% mutate(year=floor_date(as.Date(as.character(year),format = "%Y"), unit = "months"))

d_mortality_long <- pivot_longer(d_mortality,cols=c("all","0_4","5_9","10_14","15_19",         
                                    "20_24","25_29","30_34","35_39","40_44",         
                                    "45_49","50_54","55_59","60_64","65_69",         
                                    "70_74","75_79","80_84","over85"),
                         names_to = "age",
                         values_to = "count") %>% 
  filter(code %in% c(1:67)) %>% mutate(year=floor_date(as.Date(as.character(year),format = "%Y"), unit = "months"))

df_pop <- mutate(pop, year=floor_date(as.Date(as.character(year),format = "%Y"), unit = "months")) %>% 
  pivot_longer(cols=c("all","0_4","5_9","10_14","15_19",         
                      "20_24","25_29","30_34","35_39","40_44",         
                      "45_49","50_54","55_59","60_64","65_69",         
                      "70_74","75_79","80_84","over85"),
               names_to = "age",
               values_to = "population")
```

Before prediction, consider who are likely to have cancer.  
Generally, older people are vulnerable to many disease than younger.  
Take a look at time series plot by age segmentation for all cancers.  

```{r warning=F, message=F}
g1 <- filter(d_incidence_long, age != "all" & code ==1) %>%  
  ggplot() + geom_line(aes(x=year, y=count,col=age)) + labs(title = "Incidence")
g2 <- filter(d_mortality_long, age != "all" & code ==1) %>% 
  ggplot() + geom_line(aes(x=year, y=count,col=age)) + labs(title = "Mortality")
g1 + g2
```

It seems that people in older age, which is over 70 are high both in incidence and mortality counts. We are focusing on older range than 70.  
And it is quite natural that both incidents and mortality counts have increased because Japanese population is Aging year by year. We should take into account population of old age at each time.  
We use incidents and mortality counts per 1,000 over 70.  

```{r warning=F, message=F}
d_incidence_old <- filter(d_incidence_long, age %in% c("70_74","75_79","80_84","over85")) %>% 
  group_by(code,part,year) %>% summarise(count = sum(count))
d_mortality_old <- filter(d_mortality_long, age %in% c("70_74","75_79","80_84","over85")) %>% 
  group_by(code,part,year) %>% summarise(count = sum(count))
df_pop_old <- filter(df_pop, age %in% c("70_74","75_79","80_84","over85")) %>%
  group_by(year) %>% summarise(population=sum(population))

# adjust population and cancer counts
df_incidence <- left_join(d_incidence_old,df_pop_old,by="year") %>% 
  mutate(adj_count=count/population*1000)
df_mortality <- left_join(d_mortality_old,df_pop_old,by="year") %>% 
  mutate(adj_count=count/population*1000)
```

We have 28 different types of cancer data. Here we use only 4 parts; liver, bile duct, lymphoma and large intestine in order to make prediction simpler.  

**For information, liver and bile duct cancer are decreasing wheres lymphoma and large intestine are increasing.**  

### Cancer Incidence

Make time series data by indications from 1975 to 2015  

```{r warning=F, message=F}
df_incidence_liver <- ts(df_incidence[df_incidence$code==8,"adj_count"],start = 1975, frequency = 1)
df_incidence_bil <- ts(df_incidence[df_incidence$code==9,"adj_count"],start = 1975, frequency = 1)
df_incidence_lymph <- ts(df_incidence[df_incidence$code==25,"adj_count"],start = 1975, frequency = 1)
df_incidence_lintes <- ts(df_incidence[df_incidence$code==67,"adj_count"],start = 1975, frequency = 1)
```


```{r warning=F, message=F}
ggtsdisplay(df_incidence_liver, main = "Liver")
```

```{r warning=F, message=F}
ggtsdisplay(df_incidence_lymph, main = "lymphoma")
```

```{r warning=F, message=F}
ggtsdisplay(df_incidence_lintes, main = "intestine")
```

```{r warning=F, message=F}
ggtsdisplay(df_incidence_bil, main = "bile duct")
```

Considering plots above, auto correlation could be problematic.  
Usually, this kind of correlation can be better by lagged.  

```{r warning=F, message=F}
ggtsdisplay(diff(df_incidence_liver), main = "diff_Liver")
```

```{r warning=F, message=F}
ggtsdisplay(diff(df_incidence_lymph), main = "diff_limphoma")
```

```{r warning=F, message=F}
ggtsdisplay(diff(df_incidence_lintes), main = "diff_intestine")
```

```{r warning=F, message=F}
ggtsdisplay(diff(df_incidence_bil), main = "diff_bile duct")
```

By lagging 1 year, time series change gets more stationary. In this case, ARIMA model might be suitable.  
Here, train models by using data between 1975 to 2010, then predict 2011 to 2015 and compare behaviors between actual and predicted data.  

```{r warning=F, message=F}
# make training data
df_incidence_train <- filter(df_incidence, year<"2011-01-01")
df_incidence_train_liver <- ts(df_incidence_train[df_incidence_train$code==8,"adj_count"],start = 1975, frequency = 1)
df_incidence_train_lymph <- ts(df_incidence_train[df_incidence_train$code==25,"adj_count"],start = 1975, frequency = 1)
df_incidence_train_lintes <- ts(df_incidence_train[df_incidence_train$code==67,"adj_count"],start = 1975, frequency = 1)
df_incidence_train_bil <- ts(df_incidence_train[df_incidence_train$code==9,"adj_count"],start = 1975, frequency = 1)

# train models
train_incidence.liver <- auto.arima(df_incidence_train_liver,ic="aic",stepwise = F,approximation = F)
train_incidence.lymph <- auto.arima(df_incidence_train_lymph,ic="aic",stepwise = F,approximation = F)
train_incidence.lintes <- auto.arima(df_incidence_train_lintes,ic="aic",stepwise = F,approximation = F)
train_incidence.bil <- auto.arima(df_incidence_train_bil,ic="aic",stepwise = F,approximation = F)

# make plots
p_incidence_liver <- autoplot(df_incidence_liver) + autolayer(train_incidence.liver$fitted) + autolayer(forecast(train_incidence.liver,h=6),series = "ARIMA",alpha=.5) + labs(title = "Liver")
p_incidence_lymph <- autoplot(df_incidence_lymph) + autolayer(train_incidence.lymph$fitted) + autolayer(forecast(train_incidence.lymph,h=6),series = "ARIMA",alpha=.5) + labs(title = "Lymphoma")
p_incidence_lintes <- autoplot(df_incidence_lintes) + autolayer(train_incidence.lintes$fitted) + autolayer(forecast(train_incidence.lintes,h=6),series = "ARIMA",alpha=.5) + labs(title = "Intestine")
p_incidence_bil <- autoplot(df_incidence_bil) + autolayer(train_incidence.bil$fitted) + autolayer(forecast(train_incidence.bil,h=6),series = "ARIMA",alpha=.5) + labs(title = "Bile duct")
```

```{r warning=F, message=F, echo=FALSE}
p_incidence_liver
```

```{r warning=F, message=F, echo=FALSE}
p_incidence_lymph
```

```{r warning=F, message=F, echo=FALSE}
p_incidence_lintes
```

```{r warning=F, message=F, echo=FALSE}
p_incidence_bil
```

Lymphoma and Intestine (Large Intestine) are fairly well predicted while liver and bile duct are not. Especially, predicted liver moves against actual data. around 2010, some breakthrough might have happened then a number of patients ends up with decreasing.  

### Cancer Mortality

Make time series data by indications from 1958 to 2020  

```{r warning=F, message=F}
df_mortality_liver <- ts(df_mortality[df_mortality$code==8,"adj_count"],start = 1958, frequency = 1)
df_mortality_bil <- ts(df_mortality[df_mortality$code==9,"adj_count"],start = 1958, frequency = 1)
df_mortality_lymph <- ts(df_mortality[df_mortality$code==25,"adj_count"],start = 1958, frequency = 1)
df_mortality_lintes <- ts(df_mortality[df_mortality$code==67,"adj_count"],start = 1958, frequency = 1)
```


```{r warning=F, message=F}
ggtsdisplay(df_mortality_liver, main = "Liver")
```

```{r warning=F, message=F}
ggtsdisplay(df_mortality_lymph, main = "lymphoma")
```

```{r warning=F, message=F}
ggtsdisplay(df_mortality_lintes, main = "intestine")
```

```{r warning=F, message=F}
ggtsdisplay(df_mortality_bil, main = "bile duct")
```

Considering plots above, auto correlation could be problematic as well as incident data was.
So, get them lagged by a year.

```{r warning=F, message=F}
ggtsdisplay(diff(df_mortality_liver), main = "diff_Liver")
```

```{r warning=F, message=F}
ggtsdisplay(diff(df_mortality_lymph), main = "diff_limphoma")
```

```{r warning=F, message=F}
ggtsdisplay(diff(df_mortality_lintes), main = "diff_intestine")
```

```{r warning=F, message=F}
ggtsdisplay(diff(df_mortality_bil), main = "diff_bile duct")
```

By lagging 1 year, time series change gets a little more stationary but not drastically. In this case, ARIMA model might be suitable but lag can be larger than incidence.  
Here, train models by using data between 1958 to 2009, then predict 2010 to 2020 and compare behaviors between actual and predicted data.  

```{r warning=F, message=F}
# make training data
df_mortality_train <- filter(df_mortality, year<"2010-01-01")
df_mortality_train_liver <- ts(df_mortality_train[df_mortality_train$code==8,"adj_count"],start = 1958, frequency = 1)
df_mortality_train_lymph <- ts(df_mortality_train[df_mortality_train$code==25,"adj_count"],start = 1958, frequency = 1)
df_mortality_train_lintes <- ts(df_mortality_train[df_mortality_train$code==67,"adj_count"],start = 1958, frequency = 1)
df_mortality_train_bil <- ts(df_mortality_train[df_mortality_train$code==9,"adj_count"],start = 1958, frequency = 1)

# train models
train_mortality.liver <- auto.arima(df_mortality_train_liver,ic="aic",stepwise = F,approximation = F)
train_mortality.lymph <- auto.arima(df_mortality_train_lymph,ic="aic",stepwise = F,approximation = F)
train_mortality.lintes <- auto.arima(df_mortality_train_lintes,ic="aic",stepwise = F,approximation = F)
train_mortality.bil <- auto.arima(df_mortality_train_bil,ic="aic",stepwise = F,approximation = F)

# make plots
p_mortality_liver <- autoplot(df_mortality_liver) + autolayer(train_mortality.liver$fitted) + autolayer(forecast(train_mortality.liver,h=11),series = "ARIMA",alpha=.5) + labs(title = "Liver")
p_mortality_lymph <- autoplot(df_mortality_lymph) + autolayer(train_mortality.lymph$fitted) + autolayer(forecast(train_mortality.lymph,h=11),series = "ARIMA",alpha=.5) + labs(title = "Lymphoma")
p_mortality_lintes <- autoplot(df_mortality_lintes) + autolayer(train_mortality.lintes$fitted) + autolayer(forecast(train_mortality.lintes,h=11),series = "ARIMA",alpha=.5) + labs(title = "Intestine")
p_mortality_bil <- autoplot(df_mortality_bil) + autolayer(train_mortality.bil$fitted) + autolayer(forecast(train_mortality.bil,h=11),series = "ARIMA",alpha=.5) + labs(title = "Bile duct")

```

```{r warning=F, message=F, echo=FALSE}
p_mortality_liver
```

```{r warning=F, message=F, echo=FALSE}
p_mortality_lymph
```

```{r warning=F, message=F, echo=FALSE}
p_mortality_lintes
```

```{r warning=F, message=F, echo=FALSE}
p_mortality_bil
```

Lymphoma is well predicted. Liver and bile duct are predicted to be decreasing, but the gradients of them are steeper in actual data than in predicted data. Unfortunately, actual data of large intestine is higher than predicted data. This fact might indicate that the case is likely to be worse than we would consider it to be.  

## Conclusion

In this analysis, we just used time series data and did not consider environmental factor such as new drugs, medical breakthrough at all. But from these results, we can discuss and look into background.  
For example, it was a good tendency for us that liver and bile duct mortality rates were lower than predicted because it might mean that life science technology has been advanced faster than we thought. However, when it comes to large intestine, mortality rate is higher than prediction, which might mean that our diet is getting unhealthier.  
